---
title: "01_interoperability"
author: "Ronald Wu"
date: '2023-08-18'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Library installation

There was a major update to SPATA2 from version 0.1.0. Let's upgrade it to the current version 2.0.4.

We will also download the MILOLab spatial transcriptomic database package `SPATAData` to help with ST dataset download. Upgrading version from 0.0.0.9000 to new version:

```{r install libraries, eval = FALSE}
install.packages("devtools")

if (!base::requireNamespace("BiocManager", quietly = TRUE)){
      install.packages("BiocManager")
  }

BiocManager::install(c('BiocGenerics', 'DelayedArray', 'DelayedMatrixStats',
                       'limma', 'S4Vectors', 'SingleCellExperiment',
                       'SummarizedExperiment', 'batchelor', 'Matrix.utils', 'EBImage'))

install.packages("Seurat")

devtools::install_github(repo = "kueckelj/confuns")
devtools::install_github(repo = "theMILOlab/SPATAData")
devtools::install_github(repo = "theMILOlab/SPATA2")

# if you want to use monocle3 related wrappers 
devtools::install_github('cole-trapnell-lab/leidenbase')
devtools::install_github('cole-trapnell-lab/monocle3')

# SpatialExperiment

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("SpatialExperiment")
```

# Load Libraries

```{r load libraries, warning=FALSE}
library(SPATA2)
library(SPATAData)

library(Seurat)
library(tidyverse)

library(SpatialExperiment)
```

# Download datasets

## 1. 10x Genomics in-house GBM Visium sample

10X genomics houses a sample GBM Visium dataset [here](https://www.10xgenomics.com/resources/datasets/human-glioblastoma-whole-transcriptome-analysis-1-standard-1-2-0).

We will download this dataset direct from the 10x genomics website:

```{bash make folder, eval = FALSE}
cd "/Users/wuronald/WKLab Dropbox/Ronald Wu/_wouters lab/Data Ronald/spatial-transcriptomics/data"
mkdir -p data/10xgenomics/GBM/
```

```{bash download dataset, eval = FALSE}
# Input Files

wget https://cf.10xgenomics.com/samples/spatial-exp/1.2.0/Parent_Visium_Human_Glioblastoma/Parent_Visium_Human_Glioblastoma_image.tif -P /data/10xgenomics/GBM

# Output Files

wget https://cf.10xgenomics.com/samples/spatial-exp/1.2.0/Parent_Visium_Human_Glioblastoma/Parent_Visium_Human_Glioblastoma_molecule_info.h5 -P /data/10xgenomics/GBM

wget https://cf.10xgenomics.com/samples/spatial-exp/1.2.0/Parent_Visium_Human_Glioblastoma/Parent_Visium_Human_Glioblastoma_filtered_feature_bc_matrix.h5 -P /data/10xgenomics/GBM

wget https://cf.10xgenomics.com/samples/spatial-exp/1.2.0/Parent_Visium_Human_Glioblastoma/Parent_Visium_Human_Glioblastoma_filtered_feature_bc_matrix.tar.gz -P /data/10xgenomics/GBM

wget https://cf.10xgenomics.com/samples/spatial-exp/1.2.0/Parent_Visium_Human_Glioblastoma/Parent_Visium_Human_Glioblastoma_raw_feature_bc_matrix.h5 -P /data/10xgenomics/GBM

wget https://cf.10xgenomics.com/samples/spatial-exp/1.2.0/Parent_Visium_Human_Glioblastoma/Parent_Visium_Human_Glioblastoma_raw_feature_bc_matrix.tar.gz -P /data/10xgenomics/GBM

wget https://cf.10xgenomics.com/samples/spatial-exp/1.2.0/Parent_Visium_Human_Glioblastoma/Parent_Visium_Human_Glioblastoma_analysis.tar.gz -P /data/10xgenomics/GBM

wget https://cf.10xgenomics.com/samples/spatial-exp/1.2.0/Parent_Visium_Human_Glioblastoma/Parent_Visium_Human_Glioblastoma_spatial.tar.gz -P /data/10xgenomics/GBM

wget https://cf.10xgenomics.com/samples/spatial-exp/1.2.0/Parent_Visium_Human_Glioblastoma/Parent_Visium_Human_Glioblastoma_metrics_summary.csv -P /data/10xgenomics/GBM

wget https://cf.10xgenomics.com/samples/spatial-exp/1.2.0/Parent_Visium_Human_Glioblastoma/Parent_Visium_Human_Glioblastoma_web_summary.html -P /data/10xgenomics/GBM

wget https://cf.10xgenomics.com/samples/spatial-exp/1.2.0/Parent_Visium_Human_Glioblastoma/Parent_Visium_Human_Glioblastoma_cloupe.cloupe -P /data/10xgenomics/GBM

```

## 2. Ravi et al Visium GBM sample (275_T):

Using the `SPATAData` library, we will download one GBM patient sample with the ID 275_T. There are two options to load the 275_T GBM ST sample:

-   First, is to download the SPATA2 object directly: `275_T.RDS`. This is done through the function SPATAData::downloadSpataObjects(), which saves the object locally after downloading it from dropbox. I have not gotten this to work due to a `cannot read from connection error`.

-   Second, we can actually download the raw files.

```{r, eval = FALSE}

# download the object from dropbox
downloadSpataObjects(sample_names = c("275_T"), folder = "objects")
downloadSpataObjects(sample_names = c("313_T"), folder = "objects")

# download the object interactively via shinyapp
launchSpataData()
```

Alternatively, we can use the dropbox link provided as part of the error message and download it via a web browser.

```{r MILO lab metadata, eval = TRUE}
# Meta data and links for MILO Lab database
SPATAData::source_df

# specify patient 275_T:
SPATAData::source_df %>%
  filter(sample == "275_T")

# dropbox object link:
dropbox_obj_link <-SPATAData::source_df %>%
                  filter(sample == "275_T") %>%
                  select("link_spata") %>% 
                  pull
# dropbox raw files link:
dropbox_raw_link <-SPATAData::source_df %>%
                  filter(sample == "275_T") %>%
                  select("link_raw") %>% 
                  pull
```

The following are the dropbox links for the 275_T sample:


1. [RAW files](`r dropbox_raw_link`)

2. [SPATA2 object](`r dropbox_obj_link`)

# Loading data:

## 1. Load as SPATA2 Object

## 2. Load as SpatialExperiment (spe) object

Code below is based on section 2.2 spot-based SpatialExperiment vignette located [here](https://bioconductor.org/packages/release/bioc/vignettes/SpatialExperiment/inst/doc/SpatialExperiment.html#11_Introduction) with minor modifications.

```{r load as SpatialExperiment, eval = FALSE}
library(SpatialExperiment)
library(here)

# read in counts
dir <- here("data","10xgenomics","GBM") # location of the 10xgenomics GBM raw data

fnm <- here(dir,"raw_feature_bc_matrix")

sce <- DropletUtils::read10xCounts(fnm)

# read in image data
img <- readImgData(
    path = file.path(dir, "spatial"),
    sample_id = "Parent_Visium_Human_Glioblastoma")

# read in spatial coordinates
fnm <- here(dir, "spatial", "tissue_positions_list.csv")
xyz <- read.csv(fnm, header = FALSE,
    col.names = c(
        "barcode", "in_tissue", "array_row", "array_col",
        "pxl_row_in_fullres", "pxl_col_in_fullres"))

# construct observation & feature metadata
rd <- S4Vectors::DataFrame(
    symbol = rowData(sce)$Symbol)

# Can also use read10xVisium()

# construct 'SpatialExperiment'

spe <- SpatialExperiment(
    assays = list(counts = assay(sce)),
    rowData = rd, 
    colData = DataFrame(xyz), 
    spatialCoordsNames = c("pxl_col_in_fullres", "pxl_row_in_fullres"),
    imgData = img,
    sample_id = "Parent_Visium_Human_Glioblastoma")

```

